### 2021-04-27 papers/thesis/full/ch2-spherical-harmonics/spherical-harmonics.tex
Addressed more of SO comments

Jacobi operators that correspond to multiplication of the orthogonal polynomial basis, in this case the spherical harmonics, by our cartesian coordinates $x, y, z$. We can obtain the entries to these operator matrices by determining the coefficients of $x\,\Ylm(x,y,z)$, $y\,\Ylm(x,y,z)$, and $z\,\Ylm(x,y,z)$ in terms of $Y^{m'}_{l'}(x,y,z)$ for any point $(x,y,z)$ on the unit sphere.

First though, we require a few results concerning the incrementing and decrementing of the parameters for the Jacobi polynomials, and the three-term recurrence for the Jacobi polynomials, that will also prove useful later on in this chapter.
\begin{lemma}
	Let $\alpha, \beta \in \R$ be general parameters. Let $z \in [-1,1]$ and define $P_{-2}^{(\alpha+1, \beta+1)}(z) \equiv P_{-1}^{(\alpha+1, \beta+1)}(z) \equiv 0$. Then, for any $n \in \N_0$:
\bseqn{
	P_{n}^{(\alpha, \beta)}(z) =
		{1 \over 2n + \alpha + \beta + 1} \Big\{ & - {(n+\alpha)(n+\beta) \over 2n+\alpha+\beta} \: P_{n-2}^{(\alpha+1, \beta+1)}(z) \\
									  & + (n+\alpha+\beta+1) \Big[{n+\alpha \over 2n+\alpha+\beta} - {n+\beta+1 \over 2n+\alpha+\beta+2}\Big] \: P_{n-1}^{(\alpha+1, \beta+1)}(z) \\
									  & {(n+\alpha+\beta+1)(n+\alpha+\beta+2) \over 2n+\alpha+\beta+2} \: P_{n}^{(\alpha+1, \beta+1)}(z) \Big\}, \\
	(1-z^2) \, P_{n}^{(\alpha, \beta)}(z) =
		{4 \over 2n + \alpha + \beta + 1} \Big\{ & + {(n+\alpha)(n+\beta) \over 2n+\alpha+\beta} \: P_{n}^{(\alpha-1, \beta-1)}(z) \\
									  & + (n+1) \Big[{n+\alpha+1 \over 2n+\alpha+\beta+2} - {n+\beta \over 2n+\alpha+\beta}\Big] \: P_{n+1}^{(\alpha-1, \beta-1)}(z) \\
									  & - {(n+1)(n+2) \over 2n+\alpha+\beta+2} \: P_{n+2}^{(\alpha-1, \beta-1)}(z) \Big\}, \\
	z \, P_{n}^{(\alpha, \beta)}(z) = 
		& {2(n+\alpha)(n+\beta) \over (2n + \alpha + \beta)(2n + \alpha + \beta + 1)} \: P_{n-1}^{(\alpha, \beta)}(z) \\
		& {(\beta^2 - \alpha^2) \over (2n + \alpha + \beta)(2n + \alpha + \beta + 2)} \: P_{n}^{(\alpha, \beta)}(z) \\
		& {2(n+1)(n + \alpha + \beta + 1) \over 2(2n + \alpha + \beta + 1)(2n + \alpha + \beta + 2)} \: P_{n+1}^{(\alpha, \beta)}(z).
}
\end{lemma}
\begin{proof}
	The first two results are simply consequences of the relationships established for incrementing and decrementing the parameters of Jacobi polynomials detailed in \cite[18.9.5, 18.9.6]{DLMF}, along with the symmetry relation that $P_{n}^{(\alpha, \beta)}(-z) \equiv (-1)^n P_{n}^{(\beta, \alpha)}(z)$. The final one is simply a rearranging of the established three-term recurrence detailed in \cite[18.9.1, 18.9.2]{DLMF}
\end{proof}
We thus can present a useful corollary specific for our needs.
\begin{corollary}\label{corollary:jacobirelations}
	Let $l, m \in \N_0$ s.t. $0 \le m \le l$ and let $z \in [-1,1]$. Then:
\bseqn{
	(1-z^2) \, \jaclm{l}{m}(z) &=  \tilde \alpha_{l,m,1} P^{(m-1,m-1)}_{l-m}(z) + \tilde \alpha_{l,m,3} P^{(m-1,m-1)}_{l-m+2}(z) \\
	\jaclm{l}{m}(z) &= \tilde \alpha_{l,m,2} P^{(m+1,m+1)}_{l-m-2}(z) + \tilde \alpha_{l,m,4} P^{(m+1,m+1)}_{l-m}(z), \\
	z \, \jaclm{l}{m}(z) &= \tilde \gamma_{l,m,1} P^{(m,m)}_{l-m-1}(z) + \tilde \gamma_{l,m,2} P^{(m,m)}_{l-m+1}(z),
}
where
\bseqn{
	\tilde \alpha_{l,m,1} &:= {2l \over 2l+1} \\
	\tilde \alpha_{l,m,2} &:= 
		\begin{cases}
			- {l \over 2(2l+1)} \quad \text{if } l - m \ge 2 \\
			\quad \quad 0 \quad \quad \text{otherwise}
		\end{cases} \\
	\tilde \alpha_{l,m,3} &:= - {2(l-m+2)(l-m+1) \over (2l+1)(l+1)} \\
	\tilde \alpha_{l,m,4} &:= {(l+m+2)(l+m+1) \over 2(2l+1)(l+1)} \\
	\tilde \gamma_{l,m,1} &:= \begin{cases}
			\frac{l}{2l+1} \quad \text{if } l - m \ge 1 \\
			\quad 0 \quad \quad \text{otherwise}
		\end{cases} \\
	\tilde \gamma_{l,m,2} &:= \frac{(l-m+1)(l+m+1)}{(2l+1)(l+1)}.
}
\end{corollary}

\bsrefcor{corollary:jacobirelations} allows us to easily determine the aforementioned coefficients, that we will formalise in the following Lemma.

\begin{lemma}\label{lemma:shxyzrelations}
	For $l \in \No$, $m \in \Z$ s.t. $0 \le |m| \le l$, the spherical harmonics satisfy the relationships:
\bseqnnumber{
	x\,\Ylm(x,y,z) &= \alpha_{l,m,1} Y^{m-1}_{l-1}(x,y,z) +  \alpha_{l,m,2} Y^{m+1}_{l-1}(x,y,z) \nonumber \\
		     & \quad \quad \quad + \alpha_{l,m,3} Y^{m-1}_{l+1}(x,y,z) + \alpha_{l,m,4} Y^{m+1}_{l+1}(x,y,z), \label{eqn:shrecx} \\
	y\,\Ylm(x,y,z) &= \beta_{l,m,1} Y^{m-1}_{l-1}(x,y,z) +  \beta_{l,m,2} Y^{m+1}_{l-1}(x,y,z) \nonumber \\
		     & \quad \quad \quad + \beta_{l,m,3} Y^{m-1}_{l+1}(x,y,z) + \beta_{l,m,4} Y^{m+1}_{l+1}(x,y,z), \label{eqn:shrecy} \\
	z\,\Ylm(x,y,z) &= \gamma_{l,m,1} Y^{m}_{l-1}(x,y,z) + \gamma_{l,m,2} Y^{m}_{l+1}(x,y,z), \label{eqn:shrecz}
}
where
\bseqn{
	\alpha_{l,m,1} &:= 
		\begin{cases}
			\frac{\clm}{2c^{m-1}_{l-1}} \tilde \alpha_{l,m,1} \quad \text{if } m>0 \\
			\frac{\clm}{2c^{m-1}_{l-1}} \tilde \alpha_{l,|m|,2} \quad \text{if } m\le0 , \, l - |m| \ge 2 \\
			\quad \quad 0 \quad \quad \quad \quad \text{otherwise} 
		\end{cases} \\ 
	\alpha_{l,m,2} &:= 
		\begin{cases}
			\frac{\clm}{2c^{m+1}_{l-1}} \tilde \alpha_{l,m,2} \quad \text{if } m\ge0, \, l - |m| \ge 2 \\
			\frac{\clm}{2c^{m+1}_{l-1}} \tilde \alpha_{l,|m|,1} \quad \text{if } m<0 \\
			\quad \quad 0 \quad \quad \quad \quad \text{otherwise} 
		\end{cases} \\
	\alpha_{l,m,3} &:= 
		\begin{cases}
			\frac{\clm}{2c^{m-1}_{l+1}} \tilde \alpha_{l,m,3} \quad \text{if } m>0 \\
			\frac{\clm}{2c^{m-1}_{l+1}} \tilde \alpha_{l,|m|,4} \quad \text{if } m\le0 
		\end{cases} \\
	\alpha_{l,m,4} &:=
		\begin{cases}
			\frac{\clm}{2c^{m+1}_{l+1}} \tilde \alpha_{l,m,4} \quad \text{if } m\ge0 \\
			\frac{\clm}{2c^{m+1}_{l+1}} \tilde \alpha_{l,|m|,3} \quad \text{if } m<0 
		\end{cases} \\
	\beta_{l,m,j} &:= (-1)^{j+1} \: i \: \alpha_{l,m,j}, \quad j = 1,2,3,4 \\
	\gamma_{l,m,1} &:= \frac{\clm}{c^{m}_{l-1}} \: \tilde \gamma_{l,m,1} \\
	\gamma_{l,m,2} &:= \frac{\clm}{c^{m}_{l+1}} \: \tilde \gamma_{l,m,2}.
}
\end{lemma}

The results in \bsreflemma{lemma:shxyzrelations} are fairly involved, and can be derived via \textit{Clebsch-Gordan coefficients}.

@@

Three-term recurrence relations for orthogonal polynomials are well established (e.g. \cite[18.9]{DLMF}, \cite{dunkl2014orthogonal}). In a similar vein, we can obtain a recurrence relation for the spherical harmonics by combining each system in (\ref{eqn:jacobimatsdef}).
\bseqn{
	\renewcommand\arraystretch{1.3}
	\begin{pmatrix}
		1 & & & \\
		B_0-G_0(x,y,z) & A_0 & & \\
		C_1 & B_1-G_1(x,y,z) & \quad A_1 \quad & \\
		& C_2 & B_2 - G_2(x,y,z)  & \ddots \\
		& & \ddots &\ddots
	\end{pmatrix}
	\SHvec(x,y,z)
	=
	\begin{pmatrix}
		Y_0 \\ 0 \\ 0 \\ 0 \\ \vdots
	\end{pmatrix},
}
where we note $Y_0^0(x,y,z) \equiv Y_0 := c_0^0 \: P_0^{(0,0)} \equiv \half {1 \over \sqrt{\pi}}$, and for each $l = 0,1,2\dots$,
\begin{align}
	A_l &:= 
		\begin{pmatrix}
			A_{x,l} \\
			A_{y,l} \\
			A_{z,l}
		\end{pmatrix} \in \C^{3(2l+1)\times(2l+3)}, \quad
	C_l := 
		\begin{pmatrix}
			C_{x,l} \\
			C_{y,l} \\
			C_{z,l}
		\end{pmatrix} \in \C^{3(2l+1)\times(2l-1)} \quad (n \ne 0), \label{eqn:clenshawmats1} \\
	B_l &:= 
		\begin{pmatrix}
			B_{x,l} \\
			B_{y,l} \\
			B_{z,l}
		\end{pmatrix} \in \C^{3(2l+1)\times(2l+1)}, \quad
	G_n(x,y) := 
		\begin{pmatrix}
			x\idmat{2l+1} \\
			y\idmat{2l+1} \\
			z\idmat{2l+1}
		\end{pmatrix}  \in \C^{3(2l+1)\times(2l+1)}. \label{eqn:clenshawmats2}
\end{align}

For each $l = 0,1,2\dots$ let $\Dlt$ be any matrix that is a left inverse of $A_l$, i.e. such that $\Dlt A_l = \idmat{2l+3}$. Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the $\Dlt$'s, we obtain a lower triangular system \cite[p78]{dunkl2014orthogonal}, which can be expanded to obtain the recurrence:
\bseqn{
	\begin{cases}
		\SHvec_{-1}(x,y,z) := 0 \\
		\SHvec_{0}(x,y,z) := Y_0 \\
		\SHvec_{l+1}(x,y,z) = -\Dlt (B_l-G_l(x,y,z)) \SHvec_l(x,y,z) - \Dlt C_l  \,\SHvec_{l-1}(x,y,z), \quad l = 0,1,2,\dots.
	\end{cases}
}

@@

%%% TANGENT SPACE ----- VECTOR SHs
\section{Vector spherical harmonics as orthogonal vectors in three variables}\label{Section:VSHs}

Since the spherical harmonics are a basis for the surface of the sphere, and the tangent space of the sphere is spanned by the gradient and perpendicular gradient of a scalar function, we have that the gradients and perpendicular gradients of the spherical harmonics 
\bseqnnumber{
	\gradYlm &:= \nabla \Ylm = \ppx{\phi} \Ylm \phivec + {1 \over \sinphi} \ppx{\theta} \Ylm \thetavec, \nonumber \\
	\gradpYlm &:= \nabla^\perp \Ylm \equiv \rvec \times \nabla \Ylm = \ppx{\phi} \Ylm \thetavec - {1 \over \sinphi} \ppx{\theta} \Ylm \phivec, \label{eqn:VSHs}
}
span the tangent space. Here, $\rvec$ is simply the outward unit normal vector to the surface of the sphere at the point $(x,y,z)$, while $\phivec, \thetavec$ are the standard spherical coordinate unit vectors, i.e. $\phivec := [\costheta \cosphi, \sintheta \cosphi, -\sinphi]^\top$, $\thetavec := [-\sintheta, \costheta, 0]^\top$. Importantly, the set $\{\gradYlm, \gradpYlm\}$ form a complete and orthogonal basis for vector valued functions in the tangent space of the sphere and are known as \textit{vector spherical harmonics (VSHs)} \cite{barrera1985vector}, orthogonal with respect to the inner product $\ip<\vec{A}, \vec{B}> = \iint_\Omega \vec{A} \cdot \vec{B}^* \: \sinphi \: \D\varphi \: \D\theta$ -- in particular, for any $l, l' \in \N$, $m, m' \in \Z$ s.t. $-l \le m \le l$, $-l' \le m' \le l'$,
\bseqn{
	\ip<\gradYlm, \gradYlmprime> &= \iint_\Omega \gradYlm \cdot {\gradYlmprime}^* \: \sinphi \: \D\varphi \: \D\theta = \delta_{l,l'} \delta_{m,m'} \: l (l+1), \\
	\ip<\gradpYlm, \gradpYlmprime> &= \iint_\Omega \gradpYlm \cdot {\gradpYlmprime}^* \: \sinphi \: \D\varphi \: \D\theta = \delta_{l,l'} \delta_{m,m'} \: l (l+1), \\
	\ip<\gradYlm, \gradpYlmprime> &= \iint_\Omega \gradpYlm \cdot {\gradpYlmprime}^* \: \sinphi \: \D\varphi \: \D\theta = 0.
}
Vector spherical harmonics have been widely used in electrostatics (e.g. \cite{barrera1985vector}), electrodynamics (e.g. \cite{carrascal1991vector}), and of course fluid dynamics including weather and climate modelling (e.g. \cite{moses1974use, fengler2005nonlinear, swarztrauber1996spectral}). Other definitions for VSHs are used (e.g. \cite{hill1954theory}), however it is convenient for deriving explicit sparse relations and operators to use the one described here.

A simple calculation shows that such orthogonal vectors must still have block-tridiagonal Jacobi operators, as multiplication by $x, y$, or $z$ remains inside the ideal.

@@

	&= \iint_\Omega \cosphi \: \gradYlm \cdot {\gradYlmprime}^* \sinphi \: \D\phi \: \D\theta \nonumber \\

@@

	&\quadfour \cdot \Big( \int_{-1}^{1} \Big\{ \big[ \abs{m} z \jaclmabs{l}{m}(z) - \rho(z)^2 \jaclmabsdz{l}{m}(z) \big] \big[ \abs{m'} z \jaclmabs{l'}{m'}(z) - \rho(z)^2 \jaclmabs{l'}{m'}(z) \big] \nonumber \\
	&\quadeight + mm' \jaclmabs{l}{m}(z) \jaclmabs{l'}{m'}(z) \Big\} \: z \; \rho(z)^{\abs{m}+\abs{m'}-2} \: \D z \Big) \nonumber \\
	&= \clm \clmprime 2 \pi \delta_{m',m} \: \int_{-1}^{1} \Big\{ z \jaclmabs{l}{m} \jaclmabs{l'}{m'} \rho^{\abs{m}+\abs{m'}-2} \big(mm' + \abs{m}\abs{m'} z^2 \big) \nonumber \\
	&\quadeight \quadtwo - \abs{m} z^2 \jaclmabs{l}{m} \jaclmabsdz{l'}{m'} \rho^{\abs{m}+\abs{m'}} \nonumber \\
	&\quadeight \quadtwo - \abs{m'} z^2 \jaclmabsdz{l}{m} \jaclmabs{l'}{m'} \rho^{\abs{m}+\abs{m'}} \nonumber  \\

>>>>

Jacobi operators are matrices that correspond to multiplication of the orthogonal polynomial basis, in this case the spherical harmonics, by our cartesian coordinates $x, y, z$. We can obtain the entries to these operator matrices by determining the coefficients of $x\,\Ylm(x,y,z)$, $y\,\Ylm(x,y,z)$, and $z\,\Ylm(x,y,z)$ in terms of $Y^{m'}_{l'}(x,y,z)$ for any point $(x,y,z)$ on the unit sphere. Relations concerning the products of Spherical Harmonics have already been established, and so the expressions we desire here could just be seen as special cases of these. More specifically, the coefficients of the expansion in the SH basis of the product of two SHs are composed of Clebsch-Gordan coefficients \cite[p231]{sakurai1995modern}. Clebsch-Gordan coefficients are important in quantum mechanics and angular momentum, with tables and algorithms existing to calculate them \cite{mcnamee1964tables}. By simply choosing appropriate combinations, one could gain the desired coefficients for the expansion of each of $x\,\Ylm(x,y,z)$, $y\,\Ylm(x,y,z)$, and $z\,\Ylm(x,y,z)$.

However we reiterate our goal here is to use techniques that can generalise to spectral methods on other domains using multidimensional OPs, and thus we proceed by exploiting the three-term recurrences and other relations of the univariate OPs that make up the Spherical Harmonics as we have defined them. With that being said, we require a few results concerning the incrementing and decrementing of the parameters for the Jacobi polynomials, and the three-term recurrence for the Jacobi polynomials, that will also prove useful later on in this chapter.
\begin{lemma}
	Let $\alpha, \beta \in \R$ be general parameters. Let $z \in [-1,1]$ and define $P_{-2}^{(\alpha+1, \beta+1)}(z) \equiv P_{-1}^{(\alpha+1, \beta+1)}(z) \equiv 0$. Then, for any $n \in \N_0$ the following identities hold:
\bseqn{
	&1)\quad P_{n}^{(\alpha, \beta)}(z) = {1 \over 2n + \alpha + \beta + 1} \Big\{- {(n+\alpha)(n+\beta) \over 2n+\alpha+\beta} \: P_{n-2}^{(\alpha+1, \beta+1)}(z) \\
									  					& \quadfour \quadtwo \quad + (n+\alpha+\beta+1) \Big[{n+\alpha \over 2n+\alpha+\beta} - {n+\beta+1 \over 2n+\alpha+\beta+2}\Big] \: P_{n-1}^{(\alpha+1, \beta+1)}(z) \\
									 					& \quadfour \quadtwo \quad + {(n+\alpha+\beta+1)(n+\alpha+\beta+2) \over 2n+\alpha+\beta+2} \: P_{n}^{(\alpha+1, \beta+1)}(z) \Big\}, \\ \\
	&2)\quad (1-z^2) \, P_{n}^{(\alpha, \beta)}(z) = {4 \over 2n + \alpha + \beta + 1} \Big\{{(n+\alpha)(n+\beta) \over 2n+\alpha+\beta} \: P_{n}^{(\alpha-1, \beta-1)}(z) \\
									 							& \quadeight \quadtwo \: + (n+1) \Big[{n+\alpha+1 \over 2n+\alpha+\beta+2} - {n+\beta \over 2n+\alpha+\beta}\Big] \: P_{n+1}^{(\alpha-1, \beta-1)}(z) \\
									 							& \quadeight \quadtwo \: - {(n+1)(n+2) \over 2n+\alpha+\beta+2} \: P_{n+2}^{(\alpha-1, \beta-1)}(z) \Big\}, \\ \\
	&3)\quad z \, P_{n}^{(\alpha, \beta)}(z) = {2(n+\alpha)(n+\beta) \over (2n + \alpha + \beta)(2n + \alpha + \beta + 1)} \: P_{n-1}^{(\alpha, \beta)}(z) \\
								& \quadeight + {(\beta^2 - \alpha^2) \over (2n + \alpha + \beta)(2n + \alpha + \beta + 2)} \: P_{n}^{(\alpha, \beta)}(z) \\
								& \quadeight + {2(n+1)(n + \alpha + \beta + 1) \over 2(2n + \alpha + \beta + 1)(2n + \alpha + \beta + 2)} \: P_{n+1}^{(\alpha, \beta)}(z).
}
\end{lemma}
\begin{proof}
	The first two results are simply consequences of the relationships established for incrementing and decrementing the parameters of Jacobi polynomials detailed in \cite[18.9.5, 18.9.6]{DLMF}, along with the symmetry relation that $P_{n}^{(\alpha, \beta)}(-z) \equiv (-1)^n P_{n}^{(\beta, \alpha)}(z)$. The final one is simply a rearranging of the established three-term recurrence detailed in \cite[18.9.1, 18.9.2]{DLMF}
\end{proof}
We thus can present a useful corollary specific for our needs.
\begin{corollary}\label{corollary:jacobirelations}
	Let $l, m \in \N_0$ s.t. $0 \le m \le l$ and let $z \in [-1,1]$. Then:
\bseqn{
	(1-z^2) \, \jaclm{l}{m}(z) &=  \tilde \alpha_{l,m,1} P^{(m-1,m-1)}_{l-m}(z) + \tilde \alpha_{l,m,3} P^{(m-1,m-1)}_{l-m+2}(z) \\
	\jaclm{l}{m}(z) &= \tilde \alpha_{l,m,2} P^{(m+1,m+1)}_{l-m-2}(z) + \tilde \alpha_{l,m,4} P^{(m+1,m+1)}_{l-m}(z), \\
	z \, \jaclm{l}{m}(z) &= \tilde \gamma_{l,m,1} P^{(m,m)}_{l-m-1}(z) + \tilde \gamma_{l,m,2} P^{(m,m)}_{l-m+1}(z),
}
where
\bseqn{
	\tilde \alpha_{l,m,1} &:= {2l \over 2l+1}, \quad
	&&\tilde \alpha_{l,m,2} := 
		\begin{cases}
			- {l \over 2(2l+1)} \quad \text{if } l - m \ge 2 \\
			\quad \quad 0 \quad \quad \text{otherwise}
		\end{cases}, \\
	\tilde \alpha_{l,m,3} &:= - {2(l-m+2)(l-m+1) \over (2l+1)(l+1)}, \quad
	&&\tilde \alpha_{l,m,4} := {(l+m+2)(l+m+1) \over 2(2l+1)(l+1)}, \\
	\tilde \gamma_{l,m,1} &:= \begin{cases}
			\frac{l}{2l+1} \quad \text{if } l - m \ge 1 \\
			\quad 0 \quad \quad \text{otherwise}
		\end{cases}, \quad
	&&\tilde \gamma_{l,m,2} := \frac{(l-m+1)(l+m+1)}{(2l+1)(l+1)}.
}
\end{corollary}
\bsrefcor{corollary:jacobirelations} allows us to easily determine the aforementioned coefficients, that we will formalise in the following Lemma.

\begin{lemma}\label{lemma:shxyzrelations}
	For $l \in \No$, $m \in \Z$ s.t. $0 \le |m| \le l$, the spherical harmonics satisfy the relationships:
\bseqnnumber{
	x\,\Ylm(x,y,z) &= \alpha_{l,m,1} Y^{m-1}_{l-1}(x,y,z) +  \alpha_{l,m,2} Y^{m+1}_{l-1}(x,y,z) \nonumber \\
		     & \quad \quad \quad + \alpha_{l,m,3} Y^{m-1}_{l+1}(x,y,z) + \alpha_{l,m,4} Y^{m+1}_{l+1}(x,y,z), \label{eqn:shrecx} \\
	y\,\Ylm(x,y,z) &= \beta_{l,m,1} Y^{m-1}_{l-1}(x,y,z) +  \beta_{l,m,2} Y^{m+1}_{l-1}(x,y,z) \nonumber \\
		     & \quad \quad \quad + \beta_{l,m,3} Y^{m-1}_{l+1}(x,y,z) + \beta_{l,m,4} Y^{m+1}_{l+1}(x,y,z), \label{eqn:shrecy} \\
	z\,\Ylm(x,y,z) &= \gamma_{l,m,1} Y^{m}_{l-1}(x,y,z) + \gamma_{l,m,2} Y^{m}_{l+1}(x,y,z), \label{eqn:shrecz}
}
where
\bseqn{
	\alpha_{l,m,1} &:= {\clm \over 2c^{m-1}_{l-1}}
		\begin{cases}
			\tilde \alpha_{l,m,1} &\text{if } m>0 \\
			\tilde \alpha_{l,\abs{m},2} &\text{if } m\le0 \\
			0 &\text{otherwise} 
		\end{cases}, \quad
	&&\alpha_{l,m,2} := {\clm \over 2c^{m+1}_{l-1}}
		\begin{cases}
			\tilde \alpha_{l,m,2} &\text{if } m\ge0 \\
			\tilde \alpha_{l,\abs{m},1} &\text{if } m<0 \\
			0 &\text{otherwise} 
		\end{cases}, \\
	\alpha_{l,m,3} &:= {\clm \over 2c^{m-1}_{l+1}}
		\begin{cases}
			\tilde \alpha_{l,m,3}  &\text{if } m>0 \\
			\tilde \alpha_{l,\abs{m},4} &\text{if } m\le0 
		\end{cases}, \quad
	&&\alpha_{l,m,4} := {\clm \over 2c^{m+1}_{l+1}}
		\begin{cases}
			\tilde \alpha_{l,m,4} \quad \text{if } m\ge0 \\
			\tilde \alpha_{l,\abs{m},3} \quad \text{if } m<0 
		\end{cases}, \\
	\gamma_{l,m,1} &:= \frac{\clm}{c^{m}_{l-1}} \: \tilde \gamma_{l,m,1}, \quad
	&&\gamma_{l,m,2} := \frac{\clm}{c^{m}_{l+1}} \: \tilde \gamma_{l,m,2}, \\
	\beta_{l,m,j} &:= (-1)^{j+1} \: i \: \alpha_{l,m,j}, \quad j = 1,2,3,4.
}
\end{lemma}

\remark The results in \bsreflemma{lemma:shxyzrelations} are fairly involved and can be derived via the relationships established for products of Spherical Harmonics and Clebsch-Gordan coefficients as mentioned earlier, but we include this proof so as to generalise to other OPs later on in later chapters.

@@

Three-term recurrence relations for orthogonal polynomials are well established (see for example \cite[Section 18.9]{DLMF}), including for multidimensional OPs \cite[Section 3.2]{dunkl2014orthogonal}. In a similar vein, we can obtain a recurrence relation for the Spherical Harmonics by combining each system in \bsrefeqn{eqn:jacobimatsdef} in a specific way. It should be emphasised at this point that this is not necessarily the optimal way of computing the Spherical Harmonics; various efficient methods and algorithms have been proposed (see e.g. \cite{guy1998efficient, sloan2013efficient, fukushima2012numerical}). We present this method in order to be able to generalise it to other multidimensional OP families later on in this thesis.

Rewriting \bsrefeqn{eqn:jacobimatsdef}, we have that:
\bseqn{
	(J_x - xI)\SHvec(x,y,z) = (J_y - yI)\SHvec(x,y,z) = (J_z - zI)\SHvec(x,y,z) = \vec{0}.
}
Combining these relationships into a single system, complete with initial (degree 0) value, results in:
\bseqn{
	\renewcommand\arraystretch{1.3}
	\begin{pmatrix}
		1 & & & \\
		B_0-G_0(x,y,z) & A_0 & & \\
		C_1 & B_1-G_1(x,y,z) & \quad A_1 \quad & \\
		& C_2 & B_2 - G_2(x,y,z)  & \ddots \\
		& & \ddots &\ddots
	\end{pmatrix}
	\SHvec(x,y,z)
	=
	\begin{pmatrix}
		Y_0 \\ 0 \\ 0 \\ 0 \\ \vdots
	\end{pmatrix},
}
where we note $Y_0^0(x,y,z) \equiv Y_0 := c_0^0 \: P_0^{(0,0)} \equiv \half {1 \over \sqrt{\pi}}$, and for each $l = 0,1,2\dots$,
\bseqnnumber{
	A_l &:= 
		\begin{pmatrix}
			A_{x,l} \\
			A_{y,l} \\
			A_{z,l}
		\end{pmatrix} \in \C^{3(2l+1)\times(2l+3)}, \quad
	C_l := 
		\begin{pmatrix}
			C_{x,l} \\
			C_{y,l} \\
			C_{z,l}
		\end{pmatrix} \in \C^{3(2l+1)\times(2l-1)} \quad (n \ne 0), \label{eqn:clenshawmats1} \\
	B_l &:= 
		\begin{pmatrix}
			B_{x,l} \\
			B_{y,l} \\
			B_{z,l}
		\end{pmatrix} \in \C^{3(2l+1)\times(2l+1)}, \quad
	G_n(x,y) := 
		\begin{pmatrix}
			x\idmat{2l+1} \\
			y\idmat{2l+1} \\
			z\idmat{2l+1}
		\end{pmatrix}  \in \C^{3(2l+1)\times(2l+1)}. \label{eqn:clenshawmats2}
}

For each $l = 0,1,2\dots$ let $\Dlt$ be any matrix that is a left inverse of $A_l$, i.e. such that $\Dlt A_l = \idmat{2l+3}$. Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the $\Dlt$'s, we obtain a lower triangular system \cite[Section 3.2]{dunkl2014orthogonal}, which can be expanded to obtain the three-term recurrence:
\bseqn{
	\begin{cases}
		\SHvec_{-1}(x,y,z) := 0 \\
		\SHvec_{0}(x,y,z) := Y_0 \\
		\SHvec_{l+1}(x,y,z) = -\Dlt (B_l-G_l(x,y,z)) \SHvec_l(x,y,z) - \Dlt C_l  \,\SHvec_{l-1}(x,y,z), \quad l = 0,1,2,\dots.
	\end{cases}
}

@@

%%% TANGENT BUNDLE ----- VECTOR SHs
\section{Vector spherical harmonics as orthogonal vectors in three variables}\label{Section:VSHs}

We have established that the Spherical Harmonics can be viewed as a basis of orthonormal polynomials on the sphere. Specifically, given a function $f : \Omega \to \C$, we can write $f = \sum_{l = 0}^{N} \sum_{m = -l}^{l} f_{l,m} \; \Ylm$ where $\{f_{l,m}\}$ are coefficients and $N \in \N$ is sufficiently large. But what about vector-valued functions?
\begin{definition}\label{def:tangentbundle}
	Let $(x,y,z) \in \Omega$. We refer to the \textbf{tangent space} of the point $(x,y,z)$ as the space of all vectors that are orthogonal to the vector $\rvec := [x,y,x]^\top$. We further define the \textbf{tangent bundle} of $\Omega$ as the set of all such spaces for every point in $\Omega$.
\end{definition}
For the purposes of this discussion, we will only consider the tangent bundle, and thus wish to write down a basis for vector-valued functions with values in the tangent bundle. Fortunately, there is a simple solution known as the \textit{vector spherical harmonics (VSHs)} \cite{barrera1985vector}:
\bseqnnumber{
	\gradYlm &:= \nabla \Ylm = \ppx{\phi} \Ylm \phivec + {1 \over \sinphi} \ppx{\theta} \Ylm \thetavec, \nonumber \\
	\gradpYlm &:= \nabla^\perp \Ylm \equiv \rvec \times \nabla \Ylm = \ppx{\phi} \Ylm \thetavec - {1 \over \sinphi} \ppx{\theta} \Ylm \phivec. \label{eqn:VSHs}
}
Here, $\rvec$ is simply the outward unit normal vector to the surface of the sphere at the point $(x,y,z)$, while $\phivec, \thetavec$ are the standard spherical coordinate unit vectors, i.e. $\phivec := [\costheta \cosphi, \sintheta \cosphi, -\sinphi]^\top$, $\thetavec := [-\sintheta, \costheta, 0]^\top$. They are orthogonal with respect to the inner product $\ip<\vec{A}, \vec{B}> = \iint_\Omega \vec{A} \cdot \vec{B}^* \: \sinphi \: \D\varphi \: \D\theta$ -- in particular, for any $l, l' \in \N$, $m, m' \in \Z$ s.t. $-l \le m \le l$, $-l' \le m' \le l'$,
\bseqn{
	\ip<\gradYlm, \gradYlmprime> &= \iint_\Omega \gradYlm \cdot {\gradYlmprime}^* \: \sinphi \: \D\varphi \: \D\theta = \delta_{l,l'} \delta_{m,m'} \: l (l+1), \\
	\ip<\gradpYlm, \gradpYlmprime> &= \iint_\Omega \gradpYlm \cdot {\gradpYlmprime}^* \: \sinphi \: \D\varphi \: \D\theta = \delta_{l,l'} \delta_{m,m'} \: l (l+1), \\
	\ip<\gradYlm, \gradpYlmprime> &= \iint_\Omega \gradpYlm \cdot {\gradpYlmprime}^* \: \sinphi \: \D\varphi \: \D\theta = 0.
} 
Let $\uvec$ be a vector-valued function on $\Omega$ whose values are of the form
\bseqn{
	\uvec(x,y,z) = u_\varphi(x,y,z) \: \phivec + u_\theta(x,y,z) \: \thetavec, \quad (x,y,z) \in \Omega
}
for some scalar functions $u^\varphi, u^\theta$. Such a function $\uvec$ we refer to as a \textit{vector-valued function in the tangent bundle} of the sphere. Importantly, the set $\{\gradYlm, \gradpYlm\}$ form a complete and orthogonal basis for such functions \cite{barrera1985vector}. That is, given $\uvec$, there exist coefficients $\{u_{l,m}^{\gradYlm}, u_{l,m}^{\gradpYlm} \}$ such that for large enough $N$:
\bseqn{
	\uvec = \sum_{l = 0}^{N} \sum_{m = -l}^{l} u_{l,m}^{\gradYlm} \; \gradYlm + u_{l,m}^{\gradpYlm} \; \gradpYlm.
}
Vector spherical harmonics have been widely used in electrostatics (e.g. \cite{barrera1985vector}), electrodynamics (e.g. \cite{carrascal1991vector}), and of course fluid dynamics including weather and climate modelling (e.g. \cite{moses1974use, fengler2005nonlinear, swarztrauber1996spectral}). Other definitions for VSHs are used (e.g. \cite{hill1954theory}), however it is convenient for deriving explicit sparse relations and operators to use the ones described here. In particular, we can note that computing the gradient or divergence of a function is merely a matter of conversion between the SH and the VSH bases.

A simple calculation shows that such orthogonal vectors must still have block-tridiagonal Jacobi operators, as multiplication by $x, y$, or $z$ remains inside the ideal. Thus, we we can use the same techniques that we used for the scalar SHs to derive Jacobi matrices, function evaluation and transforms.

@@

	&= \iint_\Omega \cosphi \: \gradYlm \cdot {\gradYlmprime}^* \sinphi \: \D\varphi \: \D\theta \nonumber \\

@@

	&\quad \cdot \Big( \int_{-1}^{1} \Big\{ \big[ \abs{m} z \jaclmabs{l}{m}(z) - \rho(z)^2 \jaclmabsdz{l}{m}(z) \big] \big[ \abs{m'} z \jaclmabs{l'}{m'}(z) - \rho(z)^2 \jaclmabs{l'}{m'}(z) \big] \nonumber \\
	&\quadfour \quadtwo + mm' \jaclmabs{l}{m}(z) \jaclmabs{l'}{m'}(z) \Big\} \: z \; \rho(z)^{\abs{m}+\abs{m'}-2} \: \D z \Big) \label{eqn:lemmaVSHrecurrencesproofz1}
